### 2018-03-07 revfiles/ReviewerResponse.md

In the new version of the manuscript, Hannigan et al. have made a fair effort in addressing all reviewer's concerns. I find an improvement from the previous version. Sometimes these concerns were maybe a bit "over-justified" in the text, in detriment of style, but is fine.

I understand that the main intention of this manuscript is not the computational method, but the methodological approach as a whole and to show how could this approach be used to address interesting biological questions. In this sense, it is a nice contribution the use network theory combined with microbial ecology to generate biological hypothesis about how the interactions of hosts and viruses (bacteria and phages in this case) change in response to different variables and how vulnerable these associations may be to these changes. In terms of this, I find the work valuable and of interest for the scientific community.

However, I am still not quite convinced about the machine learning approach to define bacteria-phage interactions and I didn't find authors' replies on this respect convincing neither. My concern essentially stems from the biological base of their method, i.e., to which extent the connections between bacterial OGU and phage OGU have a biological base or are just computational artifacts.

> *We agree with these concerns and we think that the comments from the reviewer have really helped address them as described in our responses below..*

@@

A first caveat, acknowledged by authors and reviewers, is the fact that the training set is so small and using viruses of highly divergent environments. I understand authors responses to this issue, and I find them valid. However, one could expect that with such a poor training set, almost no links would be found in a highly complex community, with highly divergent bacteria and viruses from those use for training. Yet, many interactions were found. This is simply explained by the features that primarily defined the model.

Figure 1B shows that protein identity between viral and bacterial contigs are the primary reason, followed by nucleotide identity. This means that the random forest model will mostly look for protein similarity between phage and bacterial contigs, i.e., will predominantly look for prophages integrated in bacterial genomes. Do authors think that this is really a widespread feature between bacteria and their corresponding phages, that can be effectively used in real microbial communities? In a way, the model is only accounting for a limited and narrow sub-fraction of the phage and bacteria spectrum (i.e., only bacteria that are already infected with prophages of a certain viral family/genus, that are ALSO present as free viral particles in the sample).

> *This is a valuable point and deserves further clarrification in the manuscript. These communities have been previously described as consisting primarily of lysogenic phages. Therefore, while our approach could be improved by future work and more comprehensive datasets, we are addressing lysogenic phages, the most abundant component of the virome. Second, it is worth noting that while the alignment approaches were most informative, the protein family interaction approach was also informative. Because the protein family approach relies on protein-protein interactions and not detecting prophage sequences in the bacterial genomes, it is not expected to be limited to lysogenic phages. LINES 182 - 184*

@@

> *This is a fantastic addition to the manuscript. The suggestion is to evaluate the tolerance of the findings to inaccuracies in the network prediction model. The focus of the findings is to compare groups which were trated the same, which means that any false positives or negatives will equally impact both groups. Therefore it is important to see what level of noise is tolerated by inaccuracies in the model. We accomplished this by implementing the analysis suggested by the reviewer. We iteratively added noise into the network datasets to evaluate how the observed differences in groups was effected. We included the results of this validation in the manuscript text and included relevant figures. As suggested by the reviewer, this confirms the confidence in our findings. LINES 295 - 301, 316 - 319, 396 - 397, 555 - 558, Figure S13, S14*

2. Non realistic benchmark.
The authors evaluate their predictor on an ideal full-genome case. The authors explain in their answer to reviews that they do not know how to simulate a realistic dataset. One can easily simulate sequencing reads followed by an assembly step. As for the simulation of the bins, a simulation of co-occurrence is not necessary and the effect of the binning errors can be estimated by artificially removing some contigs, and adding some contaminating contig (one can take for example the standard values given in Supplementary Figure S3 of (MetaBat, Kang et al., 2015); it would be for example good to know to what ratio of incompleteness and contamination the predictor is still reliable). Only then the evaluation of the method on this simulated data becomes more realistic and closer to the actual use-case made in the following of the paper.

> *This is another valuable comment which certainly warrants further discussion and results. To the reviewers suggestion around simulations, one could definitely artificially create sequence datasets from the reference genomes and assemble them back into contigs. The issue we would like to point out however is that, although it was suggested above that "a simulation of co-occurrence is not necessary", the referenced citation and figure do use co-occurence abundance data in their validation (in fact, MetaBat stands for "Metagenome Binning with Abundance and Tetra-nucleotide frequencies"). The group created artificial contigs (their error free contigs) by randomly shredding reference genomes that were most highly represented in the MetaHIT dataset which they used for co-abundance calculations. In this way the authors could test their binning strategy using artificial contigs with co-abundance data from their MetaHIT dataset. As we mentioned in the last review iteration as well, we are unsure of how to accomplish this without a similar robust empirical dataset with well described interactions. At best we would be building a simulation on a simulation. To address the reviewer comments, and what we interpret to be the heart of the issue, we included a discussion as to how our method relates to this approach, what future work will be pursued, and what the ideal dataset and experimental setup will look like, in the discussion section.*

### 2018-03-07 revfiles/ReviewerResponse.md

To the editors of PLOS Computational Biology,

We would again like to thank the reviewers and editors for their encouraging, positive, insightful, and constructive comments. We appreciate the time of the editors and reviewers, and this feedback greatly improved our manuscript. We are resubmitting our manuscript with point-by-point responses to the reviewer comments below. We are also submitting a marked up revised manuscript with additions marked in blue and deletions marked in red. Included in the revisions are four additional figures, one additional table, updated text with further discussion, and an improved quality filtering process for the sequence set, as suggested. All of the analyses were rerun following the additional suggested quality filtering steps.

@@

In the new version of the manuscript, Hannigan et al. have made a fair effort in addressing all reviewer's concerns. I find an improvement from the previous version. Sometimes these concerns were maybe a bit "over-justified" in the text, in detriment of style, but is fine.

I understand that the main intention of this manuscript is not the computational method, but the methodological approach as a whole and to show how could this approach be used to address interesting biological questions. In this sense, it is a nice contribution the use network theory combined with microbial ecology to generate biological hypothesis about how the interactions of hosts and viruses (bacteria and phages in this case) change in response to different variables and how vulnerable these associations may be to these changes. In terms of this, I find the work valuable and of interest for the scientific community.

However, I am still not quite convinced about the machine learning approach to define bacteria-phage interactions and I didn't find authors' replies on this respect convincing neither. My concern essentially stems from the biological base of their method, i.e., to which extent the connections between bacterial OGU and phage OGU have a biological base or are just computational artifacts.

> *We agree with these concerns and we think that the comments from the reviewer have really helped address them as described in our responses below.*

@@

1. Overlooking the impact of wrong predictions in the following of the analysis.
The impact of the false predictions is never discussed and the reader have no idea if the false discovery rate of the predictor is likely to change the statistics extracted from each of the networks and how this FDR could affect or even reverse the biological conclusions of the paper. The reported median specificity and sensitivity of the predictor on a balanced dataset lead to an estimated FDR of 0.34. What is the impact of those wrong predictions on the subsequent analysis? Can those false predictions change the final conclusion of the paper? As a suggestion, one can measure the impact on the reported network statistics when randomly adding or removing edges corresponding to the fraction of estimated false negative rate and false positive rate. This would give an estimate of the confidence range on the final statistics of the network and assess if the differences between the samples are significant.
Even though this analysis is mandatory for the reported median performances of the predictor, as the authors say there are high variations in the sensitivity and specificity in the nested cross-validation iterations and a complementary analysis of the impact of the wrong predictions on the final results at different sensitivity-specificity trade-offs should also be reported. An estimate of the change in the final results can be computed by tuning the threshold applied on the prediction score (which is responsible for the trade-off sensitivity-specificity).

> *This is a fantastic addition to the manuscript. The suggestion is to evaluate the tolerance of the findings to inaccuracies in the network prediction model. The focus of the findings is to compare groups which were trated the same, which means that any false positives or negatives will equally impact both groups. Therefore it is important to see what level of noise is tolerated by inaccuracies in the model. We accomplished this by implementing the analysis suggested by the reviewer. We iteratively added noise into the network datasets to evaluate how the observed differences in groups was effected. We included the results of this validation in the manuscript text and included relevant figures. As suggested by the reviewer, this confirms the confidence in our findings. LINES 292 - 298, 313 - 316, 393 - 394, 552 - 555, Figure S13, S14*
